Pool Model Paper

### 2023-12-05 pool_model.tex
formatting of equations and text

To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\end{equation}
In the following section we demonstate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

\section{One Species Modelling}
\subsection{Three-Pool Model}

Consider for simplicity the system where only one bacteria species is present $N=1$.
The bacterial growth curves usually consists of different consecutive phases~\cite{buchanan_when_1997}.
At the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949}
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening \cmt{Add Ref}.
When the nutrition is exhausted by the growing bacteria, the stationary state of the growth curve occurres.
In this state the cultures accumulate waste products of metabolitic activity that leads to cell death of the bacteria.
At the same time dying cells release the nutritions for survivors to continue growing.
As a result the balance between the growing and dying processes leads to the constant with time bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happeng: adaplation, growth or death.
Motivated by this we divide a bacterial population into three pools (states) (\textbf{Figure \ref{fig:SchematicRep}}): 
\begin{itemize}
\item $x^1_1 (t) = L(t)$: the fraction of the population in the lag phase
\item $x^2_1 (t) = G(t)$: the fraction of the population growing and dividing
\item $x^3_1 (t) = D(t)$: the fraction of the population undergoing cell death
\end{itemize}
Here the concentrations of the pools are components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}
    \caption{{\footnotesize Schematic representation of exchange between three states or pools.}}
    \label{fig:SchematicRep}
    \end{center}
\end{figure}

To detremine the function $\mbF(\mbx, \mbm, \mbp, \mbu)$ consider the flows between different states.
The pools $G$ and $D$ are unidirectionally connected with the pool $L$. 
For now we assume for simplicity that there is no back flow between growth state $G$ and lag-state $L$. 
The suggested model can be cast in the following reactions: \cmt{Use notations x ??}
\begin{eqnarray}
L &\stackrel{\lambda}{\longrightarrow} & G\\
L &\stackrel{\mu}{\longrightarrow} & D\\
G &\stackrel{\alpha}{\longrightarrow} & 2G\\
G &\stackrel{\mu'}{\longrightarrow} & D\\
D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}

Then the \acp{ode} describing such a model can be written as 
\begin{equation}
\begin{cases}
    \dot{L} = -(\mu + \lambda) L\\
    \dot{G} = \lambda L + \alpha G - \mu' G\\
    \dot{D} = \mu  L + \mu' G- \beta D  
\end{cases}
\end{equation}
or in matrix form it shows the linear dependency
\begin{equation}
    \dot \mbx  = \begin{pmatrix}
                    -(\mu + \lambda) & 0             & 0      \\
                    \lambda          & \alpha - \mu' & 0      \\
                    \mu              & \mu'          & -\beta 
                 \end{pmatrix} \mbx .
\end{equation}

@@

Based on experimental observations, it was shown that the bacterial growth is strongly influenced by the limiting resource.
In the system where all needed nutritients are in excess except for one the total growth is linear to the concentration of this liminting nutritient~\cite{hibbing_bacterial_2010, monod_growth_1949}.
The control of the population density with respect to the available nutritions is implemented using cell communication process called quorum sensing. 
Using release of the extracellular signalling molecules (autoinducers) it is possible to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.

We would not dive into the processes of quorum sensing and try to keep it as simple as possible.
According to this, we aim to include a resource pool which is used by the growing bacteria using the second order kinetic reaction:
\begin{equation}
    G + R_0  \stackrel{\alpha_0}{\longrightarrow} 2G.
\end{equation}

Note that the pools $L$, $G$ and $D$ represent number of bacteria (or number of bacteria per unit area/volume) while $R_0$ represents an abstract resource pool. 
Let say $G=N_B/V$ and $[G]=1/V$, where $N_B$ is number of bacteria and $V$ is the volume. 
Then $R_0=N_R/V$ and $[R_0]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol). 
Then $[\alpha_0]=V/s$, $[N_t]=1/V$, and $[\alpha]=1/s$. 
For homogeneous densities it is equivalent to consider the total amount, i.e., multiplying $L$ and $G$ by the volume. 
In this case $[L]=[G]=[N_t]=1$.
For simplicity we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and diminsioneless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).
Here the full capacity of the resource pool $N_t$ was used as a scaling constant.
In this case the growth term can be simply written as $\alpha G R$ and the model is fully formulated in the following way: 
\begin{eqnarray}
    \dot \mbx  &=& \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
                                                    -(\mu + \lambda) & 0               & 0      \\
                                                    \lambda          & \alpha R - \mu' & 0      \\
                                                    \mu              & \mu'            & -\beta 
                                                \end{pmatrix} \mbx  \\
    \dot \mbm &=& \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{eqnarray}

Such model can be simplified by explointing that limitation of the resource pool also limits the total bacterial count of the system:
\begin{equation}
    R = 1 - \frac{L+G+D}{N_t}
\end{equation}
and, hence, the full system can be written as 
\begin{equation}
    \begin{cases}
        \dot{L} = -(\mu + \lambda) L\\
        \dot{G} = \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \dot{D} = \mu  L + \mu' G- \beta D 
        \label{eq:3pool_resource} 
    \end{cases}.
\end{equation}
Rewriting the equations above, we get the following non-linear system:
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu & + \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\end{equation}
Here the first term is the linear part and the second term define the quadratic behaviour of the system.

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_3pools_resource.pdf}
    \caption{{\footnotesize The time dependence of the total bacterial count $N(t) = \sum_i \sum_k \mbx_i^k = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
    All curves were calculated using the model (\ref{eq:3pool_resource}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.
    The purple curve shows significant dip in behaviour of the system where the death of the bacteria is present $\mu = 0.5$.
    Other curves show the case of no bacteria dying $\mu=0$ and represent different length of lag phases:
    short $\lambda=10^{-2}$ (red), intermediate $\lambda=10^{-3}$ (blue) and long one $\lambda=10^{-4}$ (green).
}}
    \label{fig:3pool_resource_plots}
    \end{center}
\end{figure}

\subsection{Lag-time Estimation}

Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$. 
The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients. 
If yes, the three pool model would be advantageous, if not the two pool model is sufficient with the advantages of having one parameter less.

The dynamics of the three pool model can be captures by the coupled system of ordinary differential equations (the two pool model can be achieved by setting $D\equiv 0$ and replacing $\mu$ with $\beta$):
to be solved with the initial conditions $L(0)=n_0$, $G(0)=0$, and $D(0)=0$, where $n_0$ is the initial bacterial population. 
$N_t$ denotes the maximal size of the total bacterial population due to environmental limitations (\textbf{Figure \ref{fig:3pool_resource_plots}}). 
For $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:
\begin{equation}
    \begin{cases}
        L(t) = n_0 e^{-(\mu+\lambda)t}\\
        D(t) = n_0 \mu e^{-\beta t}\frac{1-e^{-(\mu+\lambda-\beta)t}}{\mu+\lambda-\beta}
    \end{cases}.
\end{equation}
For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be put into the differential equation for $G$. 
However, for time dependent parameters, e.g., for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$. 
It is interesting to see what type of equations result for the total bacterial population $n=L+G+D$:
\begin{equation}
    \begin{cases}
        \dot{G} = \lambda L + \alpha G\left(1-\frac{n}{N_t}\right)\\
        \dot{n} = - \beta D + \alpha G\left(1-\frac{n}{N_t}\right)
    \end{cases}
\end{equation}
with the initial conditions $G(0)=0$ and $n(0)=n_0$. 
These equations have a clear interpretation, are derived from simple principles, and are different from the popular Barayni-Roberts and modified-Gompertz models. 
A notable difference is that the growth rate $\alpha$ is time independent. 
For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = \lim_{t\to\infty} D(t) = 0$. 
It follows immediately that $\lim_{t\to\infty} G(t) = N_t$. 
The equation exhibit a lag phase and an initial drop due to cell death. 
To find a substantial initial drop one needs $\beta \mu \gg \alpha \lambda$. 
The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $n_0=2G(t_L)$. 
Assuming $L\ll N_t$ we can ignore the non-linear term in the equation for $G$, which gives rise to:
\begin{eqnarray}
1&=& 2\int_0^{t_L} \lambda L(t)e^{\alpha(t_L-t)}dt\\
1  &=& 2\lambda e^{\alpha t_L}\frac{1-e^{-(\mu+\lambda+\alpha)t_L}}{\mu+\lambda+\alpha}.
\end{eqnarray}
For $\alpha > \mu+\lambda$ $t_L$ can be approximated by:
\begin{eqnarray}
t_L &\approx& -\frac{1}{\alpha}\ln\left(\frac{2\lambda}{\mu+\lambda+\alpha}\right).
\end{eqnarray}
The lag time $t_L$ diverges very slowly, logarithmically with $\lambda\to 0$.

@@

Due to its life in rather unfavourable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.
By defeinition dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\cite{kaprelyants_dormancy_1993}. 
This survival strategy can appear in response to different external stresses, e.g., nutritient deficiency, temperature stress or extreme fluctuations of environment parameters.
And after "difficult times" bacteria, also called viable but non-culturable (VBNC), can continue cell division~\cite{kell_viability_1998}.
The models including transition of bacteria to the dormant state were proposed and studied by Bär \etal~\cite{bar_modelling_2002} and Jones and Lennon~\cite{jones_dormancy_2010}.

@@

We explore now the idea that under certain conditions, as sudden change in the environment, part of the population $G$ will enter a lag-phase, i.e., there is a back-flow from $G$ to $L$. 
\begin{eqnarray}
    L &\stackrel{\lambda}{\longrightarrow} & G\\
    L &\stackrel{\mu}{\longrightarrow} & D\\
    G &\stackrel{\alpha}{\longrightarrow} & 2G\\
    G &\stackrel{\gamma(t)}{\longrightarrow} & L\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}
The rate $\gamma(t)$ depends on the rate of change of the environment.
Stronger and quicker changes will lead to a higher rate $\gamma$. 
\subsubsection{Maxwell type of stress - strain relation}

One of the reasons of the dormant state appearence can be also the temperature shift~\cite{oliver_viable_1995}.
Moreover, the rapid shift in temperature is associated with the appearence of an additional lag phase~\cite{zwietering_modeling_1994}.
To model it we propose a kind of visco-elastic Maxwell type of stress-strain relation and write down the ordinary differential equation for $\gamma$ to be:
\begin{eqnarray}
\dot{\gamma} &=& \Gamma \left |\dot{T}\right |-\delta \gamma.
\end{eqnarray}
We assumed that the direction of the temperature change does not matter, i.e., a change from $T=2$ to $T=12$ has the same effect as a change from $T=14$ to $T=4$. 
$\Gamma$ is a scaling factor and $\delta$ is the relaxation time for the temperature disturbance. 
Integration of this equation yields (with $\gamma(0)=0$):
\begin{eqnarray}
\gamma(t) &=& \Gamma \int_0^t \left |\dot{T}(t^{\prime})\right |e^{-\delta (t-t^{\prime})}dt^{\prime}.
\end{eqnarray}
After some manipulation of this equation we find:
\begin{eqnarray}
\gamma(t) &=& \Gamma\delta\left |\left[T[t]-T[0]e^{-\delta t}-\delta \int_0^t T(t^{\prime})e^{-\delta (t-t^{\prime})}dt^{\prime}\right]\right |.
\end{eqnarray}
This representation of the rate $\gamma(t)$ has the advantage that it only involves $T(t)$ and does not require the calculation of the derivative of $T$. 
If $T$ exhibits $n$ temperature jumps, one finds:
\begin{eqnarray}
\gamma(t) &=& \Gamma\sum_{i=1}^n \left |\Delta T_i \right |\theta(t-t_i)e^{-\delta(t-t_i)}.
\label{eq:gamma_tempshift}
\end{eqnarray}

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_3pools_resource_tempshift.pdf}
    \caption{{\footnotesize (a) The time dependence of the total bacterial count $N$ for the model including the back-flow to the lag phase due to temperature stres (\ref{eq:ODE_tempshift}).
    (b) The value of the corresponding back-flow rate $\gamma$ calculated according to (\ref{eq:gamma_tempshift}).
    The curves represent no temperature shift (blue), one temperature jump $\Delta T = 10$ at time $t=4$ (red) 
    and the series of different temperature shifts of value 10, 5 and 15 at the tme points 1, 3 asnd 7, respectively (green).
    The paremeters of the back-flow rate are $\Gamma=1$ and $\delta=5$.
    The other model parameters are $\alpha=2.5$, $\mu'=0.5$, $\beta=5$, $\lambda=10^{-2}$, $\mu = 0$ fe each curve.
    }}
    \label{fig:TempJump}
    \end{center}
\end{figure}

The ordinary differential equations for the pool model with environmental-shock based back-flow read:
\begin{equation}
    \begin{cases}
    \dot{L} =\gamma(t) G - (\mu + \lambda) L\\
    \dot{G} = \lambda L -\gamma(t) G + \alpha G\left(1-\frac{L+G+D}{N_t}\right)\\
    \dot{D} = \mu  L - \beta D
    \label{eq:ODE_tempshift}
    \end{cases}
\end{equation}

@@

@@

As it was mentioned above, bacteria usually coexist in large multispecies interacting communities whose growth is vastly determined by the developed survival mechanisms.
These survival mechanisms mostly aim to secure as much available nutrition which lead to competitive strategies or, if it is advantageous, cooperative ones~\cite{hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
Further we show that the TPM can be readily extended to the systems with more than one bacteria an take describe more complex relationships between species.

@@

One on the ways to descibe competition between bacteria populations is as a function of limiting resources ratio~\cite{tilman_resource_1977, smith_effects_2002}.
To descibe such a situation we consider for simplicity only one common limiting resource pool. 
The simplest pool model for bacterial species $A$ and $B$ encompassing lag-phases, growth-phases and a common resource pool is given by (no back-flow to the lag-pool):

>>>>

To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\end{equation}
In the following section we demonstate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

%
%
\section{One Species Modelling}
\subsection{Three-Pool Model}
Consider for simplicity the system where only one bacteria species is present $N=1$.
The bacterial growth curves usually consists of different consecutive phases~\cite{buchanan_when_1997}.
At the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949}
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening \cmt{Add Ref}.
When the nutrition is exhausted by the growing bacteria, the stationary state of the growth curve occurres.
In this state the cultures accumulate waste products of metabolitic activity that leads to cell death of the bacteria.
At the same time dying cells release the nutritions for survivors to continue growing.
As a result the balance between the growing and dying processes leads to the constant with time bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happeng: adaplation, growth or death.
Motivated by this we divide a bacterial population into three pools (states) (\textbf{Figure \ref{fig:SchematicRep}}): 
\begin{itemize}
\item $x^1_1 (t) = L(t)$: the fraction of the population in the lag phase
\item $x^2_1 (t) = G(t)$: the fraction of the population growing and dividing
\item $x^3_1 (t) = D(t)$: the fraction of the population undergoing cell death
\end{itemize}
Here the concentrations of the pools are components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}
    \caption{{\footnotesize Schematic representation of exchange between three states or pools.}}
    \label{fig:SchematicRep}
    \end{center}
\end{figure}
To detremine the function $\mbF(\mbx, \mbm, \mbp, \mbu)$ consider the flows between different states.
The pools $G$ and $D$ are unidirectionally connected with the pool $L$. 
For now we assume for simplicity that there is no back flow between growth state $G$ and lag-state $L$. 
The suggested model can be cast in the following reactions: \cmt{Use notations x ??}
\begin{eqnarray}
    L &\stackrel{\lambda}{\longrightarrow} & G\\
    L &\stackrel{\mu}{\longrightarrow} & D\\
    G &\stackrel{\alpha}{\longrightarrow} & 2G\\
    G &\stackrel{\mu'}{\longrightarrow} & D\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}
Then the \acp{ode} describing such a model can be written as 
\begin{equation}
\begin{cases}
    \dot{L} = -(\mu + \lambda) L\\
    \dot{G} = \lambda L + \alpha G - \mu' G\\
    \dot{D} = \mu  L + \mu' G- \beta D  
\end{cases}
\end{equation}
or in matrix form it shows the linear dependency
\begin{equation}
    \dot \mbx  = \begin{pmatrix}
        -(\mu + \lambda) & 0             & 0      \\
        \lambda          & \alpha - \mu' & 0      \\
        \mu              & \mu'          & -\beta 
    \end{pmatrix} \mbx .
\end{equation}
%
%

@@

Based on experimental observations, it was shown that the bacterial growth is strongly influenced by the limiting resource.
In the system where all needed nutritients are in excess except for one the total growth is linear to the concentration of this liminting nutritient~\cite{hibbing_bacterial_2010, monod_growth_1949}.
The control of the population density with respect to the available nutritions is implemented using cell communication process called quorum sensing. 
Using release of the extracellular signalling molecules (autoinducers) it is possible to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.
We would not dive into the processes of quorum sensing and try to keep it as simple as possible.
According to this, we aim to include a resource pool which is used by the growing bacteria using the second order kinetic reaction:
\begin{equation}
    G + R_0  \stackrel{\alpha_0}{\longrightarrow} 2G.
\end{equation}
Note that the pools $L$, $G$ and $D$ represent number of bacteria (or number of bacteria per unit area/volume) while $R_0$ represents an abstract resource pool. 
Let say $G=N_B/V$ and $[G]=1/V$, where $N_B$ is number of bacteria and $V$ is the volume. 
Then $R_0=N_R/V$ and $[R_0]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol). 
Then $[\alpha_0]=V/s$, $[N_t]=1/V$, and $[\alpha]=1/s$. 
For homogeneous densities it is equivalent to consider the total amount, i.e., multiplying $L$ and $G$ by the volume. 
In this case $[L]=[G]=[N_t]=1$.
For simplicity we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and diminsioneless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).
Here the full capacity of the resource pool $N_t$ was used as a scaling constant.
In this case the growth term can be simply written as $\alpha G R$ and the model is fully formulated in the following way: 
\begin{eqnarray}
    \dot \mbx  &=& \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\mu + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \mu' & 0      \\
        \mu              & \mu'            & -\beta 
    \end{pmatrix} \mbx\\
    \dot \mbm &=& \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{eqnarray}

Such model can be simplified by explointing that limitation of the resource pool also limits the total bacterial count of the system:
\begin{equation}
    R = 1 - \frac{L+G+D}{N_t}
\end{equation}
and, hence, the full system can be written as 
\begin{equation}
    \begin{cases}
        \dot{L} = -(\mu + \lambda) L\\
        \dot{G} = \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \dot{D} = \mu  L + \mu' G- \beta D 
        \label{eq:3pool_resource} 
    \end{cases}.
\end{equation}
Rewriting the equations above, we get the following non-linear system:
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu & + \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\end{equation}
Here the first term is the linear part and the second term define the quadratic behaviour of the system.
\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_3pools_resource.pdf}
    \caption{{
        \footnotesize The time dependence of the total bacterial count $N(t) = \sum_i \sum_k \mbx_i^k = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
        All curves were calculated using the model (\ref{eq:3pool_resource}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.
        The purple curve shows significant dip in behaviour of the system where the death of the bacteria is present $\mu = 0.5$.
        Other curves show the case of no bacteria dying $\mu=0$ and represent different length of lag phases:
        short $\lambda=10^{-2}$ (red), intermediate $\lambda=10^{-3}$ (blue) and long one $\lambda=10^{-4}$ (green).
    }}
    \label{fig:3pool_resource_plots}
    \end{center}
\end{figure}
%
%
\subsection{Lag-time Estimation}
Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$. 
The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients. 
If yes, the three pool model would be advantageous, if not the two pool model is sufficient with the advantages of having one parameter less.

The dynamics of the three pool model can be captures by the coupled system of ordinary differential equations (the two pool model can be achieved by setting $D\equiv 0$ and replacing $\mu$ with $\beta$):
to be solved with the initial conditions $L(0)=n_0$, $G(0)=0$, and $D(0)=0$, where $n_0$ is the initial bacterial population. 
$N_t$ denotes the maximal size of the total bacterial population due to environmental limitations (\textbf{Figure \ref{fig:3pool_resource_plots}}). 
For $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:
\begin{equation}
    \begin{cases}
        L(t) = n_0 e^{-(\mu+\lambda)t}\\
        D(t) = n_0 \mu e^{-\beta t}\frac{1-e^{-(\mu+\lambda-\beta)t}}{\mu+\lambda-\beta}
    \end{cases}.
\end{equation}
For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be put into the differential equation for $G$. 
However, for time dependent parameters, e.g., for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$. 
It is interesting to see what type of equations result for the total bacterial population $n=L+G+D$:
\begin{equation}
    \begin{cases}
        \dot{G} = \lambda L + \alpha G\left(1-\frac{n}{N_t}\right)\\
        \dot{n} = - \beta D + \alpha G\left(1-\frac{n}{N_t}\right)
    \end{cases}
\end{equation}
with the initial conditions $G(0)=0$ and $n(0)=n_0$. 
These equations have a clear interpretation, are derived from simple principles, and are different from the popular Barayni-Roberts and modified-Gompertz models. 
A notable difference is that the growth rate $\alpha$ is time independent. 
For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = \lim_{t\to\infty} D(t) = 0$. 
It follows immediately that $\lim_{t\to\infty} G(t) = N_t$. 
The equation exhibit a lag phase and an initial drop due to cell death. 
To find a substantial initial drop one needs $\beta \mu \gg \alpha \lambda$. 
The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $n_0=2G(t_L)$. 
Assuming $L\ll N_t$ we can ignore the non-linear term in the equation for $G$, which gives rise to:
\begin{eqnarray}
1&=& 2\int_0^{t_L} \lambda L(t)e^{\alpha(t_L-t)}dt\\
1  &=& 2\lambda e^{\alpha t_L}\frac{1-e^{-(\mu+\lambda+\alpha)t_L}}{\mu+\lambda+\alpha}.
\end{eqnarray}
For $\alpha > \mu+\lambda$ $t_L$ can be approximated by:
\begin{eqnarray}
t_L &\approx& -\frac{1}{\alpha}\ln\left(\frac{2\lambda}{\mu+\lambda+\alpha}\right).
\end{eqnarray}
The lag time $t_L$ diverges very slowly, logarithmically with $\lambda\to 0$.
%
%

@@

Due to its life in rather unfavourable conditions bacteria developed different surviving mechanisms, one of which is adapting their metabolism and transitioning into dormant state.
By defeinition dormancy in bacteria is a reversible state where cell continue living with low metabolic activity while the cell division stops~\cite{kaprelyants_dormancy_1993}. 
This survival strategy can appear in response to different external stresses, e.g., nutritient deficiency, temperature stress or extreme fluctuations of environment parameters.
And after "difficult times" bacteria, also called viable but non-culturable (VBNC), can continue cell division~\cite{kell_viability_1998}.
The models including transition of bacteria to the dormant state were proposed and studied by Bär \etal~\cite{bar_modelling_2002} and Jones and Lennon~\cite{jones_dormancy_2010}.
%

@@

%
We explore now the idea that under certain conditions, as sudden change in the environment, part of the population $G$ will enter a lag-phase, i.e., there is a back-flow from $G$ to $L$. 
\begin{eqnarray}
    L &\stackrel{\lambda}{\longrightarrow} & G\\
    L &\stackrel{\mu}{\longrightarrow} & D\\
    G &\stackrel{\alpha}{\longrightarrow} & 2G\\
    G &\stackrel{\gamma(t)}{\longrightarrow} & L\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}
The rate $\gamma(t)$ depends on the rate of change of the environment.
Stronger and quicker changes will lead to a higher rate $\gamma$. 
%
%
\subsubsection{Maxwell type of stress - strain relation}
One of the reasons of the dormant state appearence can be also the temperature shift~\cite{oliver_viable_1995}.
Moreover, the rapid shift in temperature is associated with the appearence of an additional lag phase~\cite{zwietering_modeling_1994}.
To model it we propose a kind of visco-elastic Maxwell type of stress-strain relation and write down the ordinary differential equation for $\gamma$ to be:
\begin{eqnarray}
    \dot{\gamma} &=& \Gamma \left |\dot{T}\right |-\delta \gamma.
\end{eqnarray}
We assumed that the direction of the temperature change does not matter, i.e., a change from $T=2$ to $T=12$ has the same effect as a change from $T=14$ to $T=4$. 
$\Gamma$ is a scaling factor and $\delta$ is the relaxation time for the temperature disturbance. 
Integration of this equation yields (with $\gamma(0)=0$):
\begin{eqnarray}
    \gamma(t) &=& \Gamma \int_0^t \left |\dot{T}(t^{\prime})\right |e^{-\delta (t-t^{\prime})}dt^{\prime}.
\end{eqnarray}
After some manipulation of this equation we find:
\begin{eqnarray}
    \gamma(t) &=& \Gamma\delta\left |\left[T[t]-T[0]e^{-\delta t}-\delta \int_0^t T(t^{\prime})e^{-\delta (t-t^{\prime})}dt^{\prime}\right]\right |.
\end{eqnarray}
This representation of the rate $\gamma(t)$ has the advantage that it only involves $T(t)$ and does not require the calculation of the derivative of $T$. 
If $T$ exhibits $n$ temperature jumps, one finds:
\begin{eqnarray}
    \gamma(t) &=& \Gamma\sum_{i=1}^n \left |\Delta T_i \right |\theta(t-t_i)e^{-\delta(t-t_i)}.
    \label{eq:gamma_tempshift}
\end{eqnarray}

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_3pools_resource_tempshift.pdf}
    \caption{{\footnotesize (a) The time dependence of the total bacterial count $N$ for the model including the back-flow to the lag phase due to temperature stres (\ref{eq:ODE_tempshift}).
    (b) The value of the corresponding back-flow rate $\gamma$ calculated according to (\ref{eq:gamma_tempshift}).
    The curves represent no temperature shift (blue), one temperature jump $\Delta T = 10$ at time $t=4$ (red) 
    and the series of different temperature shifts of value 10, 5 and 15 at the tme points 1, 3 asnd 7, respectively (green).
    The paremeters of the back-flow rate are $\Gamma=1$ and $\delta=5$.
    The other model parameters are $\alpha=2.5$, $\mu'=0.5$, $\beta=5$, $\lambda=10^{-2}$, $\mu = 0$ fe each curve.
    }}
    \label{fig:TempJump}
    \end{center}
\end{figure}
The ordinary differential equations for the pool model with environmental-shock based back-flow read:
\begin{equation}
    \begin{cases}
    \dot{L} =\gamma(t) G - (\mu + \lambda) L\\
    \dot{G} = \lambda L -\gamma(t) G + \alpha G\left(1-\frac{L+G+D}{N_t}\right)\\
    \dot{D} = \mu  L - \beta D
    \label{eq:ODE_tempshift}
    \end{cases}
\end{equation}

@@

%
%

@@

As it was mentioned above, bacteria usually coexist in large multispecies interacting communities whose growth is vastly determined by the developed survival mechanisms.
These survival mechanisms mostly aim to secure as much available nutrition which lead to competitive strategies or, if it is advantageous, cooperative ones~\cite{hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
Further we show that the TPM can be readily extended to the systems with more than one bacteria an take describe more complex relationships between species. 
%
%

@@

One on the ways to descibe competition between bacteria populations is as a function of limiting resources ratio~\cite{tilman_resource_1977, smith_effects_2002}.
To descibe such a situation we consider for simplicity only one common limiting resource pool. 
The simplest pool model for bacterial species $A$ and $B$ encompassing lag-phases, growth-phases and a common resource pool is given by (no back-flow to the lag-pool):
\begin{equation}
    \begin{cases}
        \dot{L}_A = - \lambda_A R L_A\\
        \dot{L}_B = - \lambda_B R L_B \\
        \dot{G}_A = \lambda_A R L_A +\alpha_A R G_A\\
        \dot{G}_B = \lambda_B R L_B +\alpha_B R G_B\\
        \dot{R} =-\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B
    \end{cases}
    \label{eq:model_2sp_resource_comp}
\end{equation}
This system of equations defines the microenvironment $\mbm$ and bacteria state matrix $\mbx$ as
\begin{equation}
    \mbm = R, \hspace{1cm}
    \mbx = \begin{pmatrix}
        L_A & L_B \\
        G_A & G_B 
    \end{pmatrix}.
\end{equation}
We again rescaled the resource pool with $N_t$ and find
\begin{eqnarray}
\label{Resource}
R &=&1-\frac{L_A+G_A+L_B+G_B}{N_t}
\end{eqnarray}
Inserting Eq. \ref{Resource} into Eqs. \ref{eq:model_2sp_resource_comp} yields:
\begin{eqnarray*}
    \dot{L}_A &=& - \lambda_A  L_A + \frac{\lambda_A}{N_t}\left(L_A+G_A+L_B+G_B\right)L_A\\
    \dot{L}_B &=& - \lambda_B L_B + \frac{\lambda_B}{N_t}\left(L_A+G_A+L_B+G_B\right)L_B \\
    \dot{G}_A &=&  \lambda_A  L_A - \frac{\lambda_A}{N_t}\left(L_A+G_A+L_B+G_B\right)L_A +\alpha_A G_A - \frac{\alpha_A}{N_t}\left(L_A+G_A+L_B+G_B\right)G_A\\
